---
title: "Cluster quality control"
author: "Nadine Bestard"
date: "28/05/2021"
output:
  html_document:
    df_print: paged
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### set-up

```{r set-up, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(here) # reproducible paths
library(scater) # plot reduced dims
library(tibble)  # for `rownames_to_column`
library(scDblFinder) # detect doublets
library(pals) # for palettes with large n #kelly()22, #polychrome()#36, unname(alphabet())
library(dplyr) # df filtering
```

```{r project}
project <- "fire-mice"
```

```{r load}
if (!file.exists(here("processed", project, "sce_anno_01.RDS"))) {
  sce <- readRDS(here("processed", project, "sce_clusters_01.RDS"))
}else{
  sce <- readRDS(here("processed", project, "sce_anno_01.RDS"))
}
```

```{r colours}
cols25 <- unname(cols25())
# remove the black and white from the pallete, and the similars to cols25
# assessed with pal.bands and pal.cube
kelly_col <- unname(kelly()[-c(1,2,5,6)])
# remove the colours that are similar to the the 
cols25 <- 
cols <- c(kelly_col, cols25)
```

### Introduction

As in [publication] we will perform a cluster QC to remove clusters of poorer
quality. This will be assessed by the number of UMI counts, the mitochondrial
percentage, doublet analysis, ribosomal genes and the number of mice that
contribute to each cluster. Moreover we will keep in mind our experimental
groups in order to ensure biological effects are not being lost.

```{r}
plotTSNE(sce, colour_by = "cluster_names", point_size=0.5,
          text_by = "cluster_names", text_size = 3) +
  scale_color_manual(values = cols) 
```

### Number of molecules per cluster

```{r}
plotTSNE(sce, colour_by = "total") + 
  ggtitle("Total number of umi counts per cell")
plotTSNE(sce, colour_by = "detected") +
  ggtitle("Detected number of genes per cell")
```

Lower values of umi counts and detected genes can be associated to lower quality
cells. Cells can also have lower expressed genes due to their biological state
or celltype.

```{r}
pct_cells<- 50
min_umi  <- 3000
```

Select clusters with `r pct_cells` % cells having less than `r min_umi`umi
counts.

```{r}
umi_df <- as.data.frame(cbind(sce$total, as.character(sce$cluster_names)))
colnames(umi_df) <- c("umi", "cluster_names")

# caculate the percentatges of cells that are "low umi"
umi_df <- 
  umi_df %>% 
  mutate(low_umi = as.numeric(umi) < min_umi) %>% 
  group_by(cluster_names) %>% 
  summarise(n_low_umi = sum(low_umi), n_cells = n()) %>%
  mutate(pct_cell_low_umi = (n_low_umi/n_cells)*100) 
umi_df %>% arrange(desc(pct_cell_low_umi))

# Filter the clusters that have a high proportion of "low umi"
low_umi_clusters <- 
  umi_df %>% 
  filter(pct_cell_low_umi > pct_cells) %>% 
  .$cluster_names
```

The clusters flagged are `r low_umi_clusters`

### Mithocondrial genes

High mithocondrial genes is associated with stressed, lower quality, cells.

```{r}
plotTSNE(sce, colour_by = "subsets_mt_percent", text_by = "cluster_names") + 
  ggtitle("Percentatge mithocondrial genes")
```

```{r}
pct_cells<- 50
pct_mt  <- 10
```

Select clusters with `r pct_cells` % cells having more than `r pct_mt`%
mithocondrial genes.

```{r}
mt_df <- as.data.frame(cbind(sce$subsets_mt_percent, as.character(sce$cluster_names)))
colnames(mt_df) <- c("subsets_mt_percent", "cluster_names")

mt_df <- 
  mt_df %>% 
  mutate(high_pct = as.numeric(subsets_mt_percent) > 10) %>% 
  group_by(cluster_names) %>% 
  summarise(n_high_mt = sum(high_pct), n_cells = n()) %>% 
  mutate(pct_cell_high_mt = (n_high_mt/n_cells)*100) %>%  
  arrange(desc(pct_cell_high_mt))
mt_df
high_mt_clusters <- 
  mt_df %>% 
  filter(pct_cell_high_mt > pct_cells) %>% 
  .$cluster_names

```

The clusters flagged are `r high_mt_clusters`.

### Ribosomal genes

To visualise the ribosomal content in the whole dataset we plotted the cells
according to their ribosomal content. High ribosomal content in one cluster,
that expresses a mix profile, could indicate that the cells are clustering based
on ribosomal gene content.

```{r}
# the cluster 
# save ribosomal genes
is_ribo <- grepl("^Rp[sl]", rownames(sce))
# recompute the quality metrics, subseting the ribosomal genes
ribo_qc <- perCellQCMetrics(sce, subsets = list(ribo = is_ribo))
# we are only interested in ribo
sce$subsets_ribo_percent <- ribo_qc$subsets_ribo_percent
sce$subsets_ribo_sum <- ribo_qc$subsets_ribo_sum
sce$subsets_ribo_detected <- ribo_qc$subsets_ribo_detected

```

```{r}
plotTSNE(sce, colour_by = "subsets_ribo_percent") + 
  ggtitle("Percentatge ribosomal genes")
```

### Number of mice per cluster

How many mice contribute to each cluster?

```{r mice-per-cluster}
# count how many cells from each gnt group  there are per cluster
sum_per_gnt_cluster <- table(sce$genotype, sce$cluster_names )
# for each cluster count how many cells are from each mice, dividing by KO and WT mice
sum_per_mice_cluster <- table(sce$mouse, sce$genotype, sce$cluster_names )
# For each cluster sum of mice that do have cells on that cluster
colSums(sum_per_mice_cluster > 0)

# create a summary 
summary <- as.data.frame(rbind(colSums(sum_per_mice_cluster > 0), sum_per_gnt_cluster, colSums(sum_per_gnt_cluster)))
row.names(summary) <- c("HET mice","KO mice", "WT mice", "HET cells","KO cells", "WT cells", "total cells")
summary
```

Except from the obvious microglia clusters, where the numbers are very low or
even absent in the fire mice nothing stands out.

### Clusters of doublet cells

Detection of clusters formed by doublets/multiplets (i.e. multiple cells
captured within the same droplet or reaction volume). The function test each
cluster against the null hypothesis that it *does* consist of doublets. The tool
consider every possible triplet of clusters consisting of a query cluster and
two putative "source" clusters. Under the null hypothesis that the query
consists of doublets from the two sources, it computes the number of genes
(`num.de`) that are differentially expressed in the same direction in the query
cluster compared to both of the source clusters. Such genes would be unique
markers for the query cluster and provide evidence against the null hypothesis.

-   Clusters with the few unique genes (`num.de`) are more likely to be composed
    of doublets.

-   Potential doublets should also have larger library size than the cells from
    the source clusters, resulting in `lib.size` ratios bellow unity.

```{r}
if(!file.exists(here("outs",project,  "doublet_clusters_res0.6.csv"))){
  res_dbl <- findDoubletClusters(sce, sce$cluster_names)
  res_dbl <- res_dbl %>% 
    as.data.frame() %>% 
    rownames_to_column("cluster")
  write.csv(res_dbl, here("outs", project,  "doublet_clusters_res0.6.csv"))
}else{
  res_dbl <- read.csv(here("outs", project,  "doublet_clusters_res0.6.csv"), row.names = 1)
}
res_dbl
```

Analyse the results:

microglia_OPCs OPCs_3 oligo_5 microglia_3 OPC_or_oligo microglia_2 and
astrocyte_OPCs are the clusterst with less `num.de`.

microglia_OPCs, microglia_3 and microglia_2: These populations could be
microglia pruning bits of astrocytes or oligodendrocytes, we do not keep these
populations in the downstream analysis.

OPCs_3: Clear doublet as it expresses markers from neurons and OPCs.

oligo_5: Potential doublet between two other clusters that are formed of
doublets too, with very low `num.de`; we do not keep it.

OPC_or_oligo: This could seem to be a transitioning state, however they do
express multiple markers usually found in late Oligodendrocytes.

astrocyte_OPCs: These cells could also be transitioning state, and the
`lib.size` is not bellow the unity, so we first did a preliminary analysis
keeping them, but reran doubletfinder after deleting the other doublet cluster
and finaly decided to be stricter and deleting them.

oligo_4: This cluster is kept, as there is supporting evidence for Astro-Oligos
populations in the mouse brain, it will be renamed as astro-oligo in the next
annotation. <https://doi.org/10.1101/2021.04.11.439393>

<details>

<summary>

Note on astrocyte-OPCs:

</summary>

We also tested keeping them and rerunning doublet finder,

                          source1       source2 num.de median.de          best     p.value lib.size1 lib.size2        prop
    OPC-Astrocyte           OPC_1   Astrocyte_2    149    3603.0         Gnao1    7.180342e-83   0.6135593 0.4465331 0.005588435            
    Oligo-Astrocyte       Oligo_1   Astrocyte_1    254    3264.0       mt-Atp6    3.848991e-35   0.8738298 0.4604424 0.013932279
    OEG             ChP_epithelia   Astrocyte_2    623    2344.0           Npy    1.935603e-29   1.5961975 0.8223610 0.001940429
    [...]

Code in commit 0b2c63cd from the github repo, in annotation_02.rmd

</details>

```{r}
doublet<-c("microglia_OPCs", "OPCs_3", "oligo_5", "microglia_3", "OPC_or_oligo", "microglia_2", "astrocyte_OPCs")
```

### Control vs fire mice

We want to have a closer look at the clusters that do have a difference between
the knockout and the wild type before deleting the clusters.

```{r fig.width=10, message=FALSE}
# divide the two objects
sce_ko <- sce[,sce$genotype == "KO"]
sce_ctr <- sce[,sce$genotype == "WT"]
# plot them side by side
gridExtra::grid.arrange(
plotTSNE(sce_ko, colour_by = "cluster_names", point_size=0.5,
         point_alpha = 0.3, text_by = "cluster_names", text_size = 3) +
  scale_color_manual(values = cols) +
  ggtitle("fire mice"),
plotTSNE(sce_ctr, colour_by = "cluster_names", point_size=0.5,
         point_alpha = 0.3, text_by = "cluster_names", text_size = 3) +
  scale_color_manual(values = cols) +
  ggtitle("control"), 
ncol = 2
)
```

#### Proportion KO-WT

In order to visualise the proportions from KO and WT for each cluster, we do not
take in consideration the microglia clusters, as these are only present in the
control, and we normalise per number of cells per cluster.

```{r}
# delete the microglia
sce_no_mc <- sce[,!(sce$cluster_names %in% c(5, 9, 25))]

# count how many cells from each gnt group  there are per cluster
sum_per_gnt_cluster_no_mc <- table(sce_no_mc$genotype, sce_no_mc$cluster_names )

# normalise per cluster, looking how the KO and WT are distributed
# across the clusters, to give to both groups the same weight
prop_per_cluster_no_mc <- prop.table(sum_per_gnt_cluster_no_mc, margin = 1)

# calculate the proportions for each cluster
prop_per_gnt_no_mc <- round(prop.table(prop_per_cluster_no_mc , margin = 2 )*100, 2)

# Display
prop_per_gnt_no_mc<- as.data.frame(prop_per_gnt_no_mc)
colnames(prop_per_gnt_no_mc) <- c("Genotype", "cluster", "Proportion")

prop_per_gnt_no_mc %>% 
  filter(Genotype == "KO") %>% 
  arrange(desc(Proportion))

```

visualise in a plot

```{r plot}
levels(prop_per_gnt_no_mc$Genotype) <- c("KO", "HET", "WT")
ggplot(data = prop_per_gnt_no_mc, aes(x = cluster, y = Proportion, fill = Genotype)) +
  geom_bar(position = "fill", stat = "identity") + theme_classic() 

```

```{r}
ko_pct <- 60
```

```{r}
difference_KO_WT <-
  prop_per_gnt_no_mc %>% 
  arrange(desc(Proportion)) %>% 
# the microglia clusters will be the top differences, but this is not 
    # something we are specially interested
  filter(!(cluster %in% c(21, 24, 1, 16))) %>% 
    filter(Genotype == "KO") %>% 
    # select the "interesting" clusters
  filter(Proportion > ko_pct ) %>% 
  .$cluster
```

```{r}
gt_ko_pct <- unique(sce$cluster_names[sce$cluster_names %in% c( high_mt_clusters, low_umi_clusters, doublet) &
    (sce$cluster_names %in% difference_KO_WT)])
```

There are `r length(gt_ko_pct)` clusters flagged as poor quality and also more
abundant in the KO

### Cluster QC

the clusters highlighted as:

-   low umi: majority of cells having less than 3000 umi counts

-   high mt : majority of cells having more than 10 % mitochondrial genes

-   doublet: clusters composed by cells that express markers from different cell
    types.

-   There was no clusters flagged from the other categories

```{r filterout, echo=TRUE}
sce$filter_out <- sce$cluster_names %in% c( high_mt_clusters, low_umi_clusters, doublet) 
filter_out <- sum(sce$filter_out)
plotTSNE(sce, colour_by = "filter_out", point_size=0.5,
         point_alpha = 0.3, text_by = "cluster_names", text_size = 3) 
plotTSNE(sce, colour_by = "filter_out", point_size=0.5,
         point_alpha = 0.3, text_by = "celltype", text_size = 3) 
if (!file.exists(here("processed", project, "sce_clusterQC.RDS"))){
 sce <- sce[, sce$filter_out == FALSE]
 saveRDS(sce, here("processed", project, "sce_clusterQC.RDS"))
}


```

This would filter `r filter_out` cells filtered out, leaving an object with
`r sum(sce$filter_out == FALSE)`cells.

## Session Info

<details>

<summary>

Click to expand

</summary>

```{r session-info}
sessionInfo()
```

</details>
