---
title: "Reanalise after QC"
author: "Nadine Bestard"
date: "17/01/2022"
output:
  html_document:
    code_folding: hide
---


# Set-up
```{r output-code, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-up, message=FALSE, warning=FALSE}
library(here) # for reproducible paths
library(SingleCellExperiment)
library(scater) # For qcs
library(scran) # for normalisation, feature selection
library(ggplot2) # To add titles to plots
library(batchelor) # Batch correction
library(patchwork) # agregate plots
library(pals) # for palettes with large n
```

```{r load-sce}

project <- "fire-mice"
sce <- readRDS(here("processed", project, "sce_clusterQC.RDS"))
source(here("src/colours.R"))
```

The object has `r dim(sce)[1]` genes and `r dim(sce)[2]` cells.

# Cell QC
Previous thresholds were:
detected genes: 429 - 7111 
sum umi: 462 - 19024
mt: 14.37%


<details>

<summary> Click to expand plots </summary>

### Violin plots

Split by sample, colour by cell type
```{r}
plotColData(sce, x = "Sample", y = "subsets_mt_percent", colour_by = "cluster_names") +
  scale_y_log10() + ggtitle("mt genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
plotColData(sce, x = "Sample", y = "detected", colour_by = "cluster_names") +
  scale_y_log10() + ggtitle("Detected genes log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
plotColData(sce, x = "Sample", y = "sum", colour_by = "cluster_names") +
  scale_y_log10() + ggtitle("Total count log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
```

Split by cell type
```{r}
plotColData(sce, x = "cluster_names", y = "subsets_mt_percent", colour_by = "Sample") +
   ggtitle("mt genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)
plotColData(sce, x = "cluster_names", y = "detected",  colour_by = "Sample") +
   ggtitle("Detected genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)

plotColData(sce, x = "cluster_names", y = "sum",  colour_by = "Sample") +
  ggtitle("Total count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)
```

Split by cell type, log scales
```{r}
plotColData(sce, x = "cluster_names", y = "detected",  colour_by = "Sample") +
  scale_y_log10() + ggtitle("Detected genes log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
plotColData(sce, x = "cluster_names", y = "sum",  colour_by = "Sample") +
  scale_y_log10() + ggtitle("Total count log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)
```


```{r}
hist(
  sce$detected,
  breaks = 100
)
hist(
  sce$sum,
  breaks = 100
)
hist(
  sce$subsets_mt_percent,
  breaks = 100
)
```


### Scatter plots

```{r}
list_sce <- lapply(unique(sce$Sample), function(X){subset(sce, , Sample = X)})
list_plots <- lapply(list_sce, function(sce){
plotColData(sce, x = "sum", y = "subsets_mt_percent", colour_by = "cluster_names")

})
wrap_plots(list_plots)

plotColData(sce, x = "sum", y = "subsets_mt_percent", colour_by = "Sample")
plotColData(sce, x = "sum", y = "detected", colour_by = "cluster_names")
plotColData(sce, x = "sum", y = "detected", colour_by = "cluster_names")
```

</details>

# Gene QC

In the preliminary analysis we only deleted genes that had less than 2 cells expressing it, now 
we filter all genes with less than 10 cells expressing it
```{r}
# at least 10 cells should express the gene
keep_feature <- rowSums(counts(sce) > 0) > 10
sce <- sce[keep_feature,]
```


# Normalisation

```{r eval=FALSE}
## Normalise by deconvolution ---
if(!file.exists(here("processed", project,  "sce_norm_02.RDS"))){
# For reproducibility
set.seed(100)
# Quick clustering to pool samples together and deal with 0 counts
quick_clusters <- quickCluster(sce)
# Calculate size factors
sce <-
  computeSumFactors(sce, cluster = quick_clusters, min.mean = 0.1)
# Check that there are not negative size factors
summary(sizeFactors(sce))
# Apply size factors and log transform them
sce <- logNormCounts(sce)
# save object
saveRDS(sce, here("processed", project,  "sce_norm_02.RDS"))
}
```

# Feature selection

After the quality control the variable genes need to be selected again to better represent the variance in this cleaned dataset.
We follow the same methods than for our first feature selection, selecting here for 
the top 2000 genes.

```{r}
gene_var_df <- modelGeneVar(sce)
gene_var <- metadata(gene_var_df)
plot(gene_var$mean, gene_var$var, xlab= "Mean of log-expression", ylab= "Variance of log-expression")
curve(gene_var$trend(x), lwd=2, add=T, col = "red")
# select hvgs
hvgs <- getTopHVGs(gene_var_df, n=2000)
# save them in the object
rowSubset(sce) <- hvgs
```
# Dimensionality reduction

## Run PCA and choose PCs
```{r}
if(!file.exists(here("processed", project,  "sce_dimred_02.RDS"))){
set.seed(1000)
sce <- runPCA(sce)

pct_var <- attr(reducedDim(sce, "PCA"), "percentVar")
plot(pct_var, log="y", xlab="PC", ylab="pct variance explained")



 #will keep 26PCs 

#keep the previous dimensional reduction just in case
reducedDim(sce, "PCA_all") <- reducedDim(sce, "PCA")
# And replace the default PCA with the reduced PCs
reducedDim(sce, "PCA") <- reducedDim(sce, "PCA")[,1:26]
}
```

## Visualisation
For visualisation, reduce to 2 dimensions. Non linear reductions.

### UMAP and TSNE
```{r}
if(!file.exists(here("processed", project,  "sce_dimred_02.RDS"))){
sce <- runUMAP(sce, dimred="PCA")
plotReducedDim(sce, dimred="UMAP", colour_by="genotype", point_size=0.1, point_alpha = 0.3)
sce <- runTSNE(sce, dimred="PCA")
plotReducedDim(sce, dimred="TSNE", colour_by="genotype", point_size=0.1, point_alpha = 0.3)
plotReducedDim(sce, dimred="TSNE", colour_by="chip", point_size=0.1, point_alpha = 0.3)

saveRDS(sce, here("processed", project, "sce_dimred_02.RDS"))

}else{ # load teh corrected object
  sce <- readRDS(here("processed", project,  "sce_corrected_02.RDS"))
  plotReducedDim(sce, "TSNE_uncorrected", colour_by="chip")
}
```

## Rerrun barch correction
```{r batch, message=FALSE, warning=FALSE}
if (!(file.exists(
  here("processed", project,  "sce_corrected_02.RDS")
))) {
set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$chip),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  merge.order = 
    list(list("3","5"), list("4","6")),
  d = 23,
  prop.k=0.10
  )
)
# recompute dimensional reduction 
#keeping  the previous dimensional reduction
reducedDim(sce, "TSNE_uncorrected") <- reducedDim(sce, "TSNE")
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")
# save
saveRDS( sce, here("processed", project,  "sce_corrected_02.RDS"))
}
plotReducedDim(sce, colour_by= "chip", dimred = "TSNE") + 
      ggtitle("TSNE dimensional reduction corrected")
```

