---
title: "Reanalise after QC"
author: "Nadine Bestard"
date: "17/01/2022"
output:
  html_document:
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

# Set-up

```{r output-code, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-up, message=FALSE, warning=FALSE, eval=TRUE }
library(here) # for reproducible paths
library(SingleCellExperiment)
library(scater) # For qcs
library(scran) # for normalisation, feature selection
library(ggplot2) # To add titles to plots
library(batchelor) # Batch correction
library(patchwork) # agregate plots
library(pals) # for palettes with large n
```

```{r load-sce, eval=TRUE}

project <- "fire-mice"
sce <- readRDS(here("processed", project, "sce_clusterQC.RDS"))
source(here("src/colours.R"))
levels(sce$genotype) <- levels(c("WT", "HET", "KO"))
```

The object has `r dim(sce)[1]` genes and `r dim(sce)[2]` cells.

# Cell QC

Previous thresholds were: 

detected genes: 429 - 7111 

sum umi: 462 -
19024 

mt: 14.37%

<details>

<summary>

Click to expand plots

</summary>

### Violin plots

Split by sample

```{r violinsample, eval=TRUE}
plotColData(sce, x = "Sample", y = "subsets_mt_percent") +
  scale_y_log10() + ggtitle("mt genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
plotColData(sce, x = "Sample", y = "detected") +
  scale_y_log10() + ggtitle("Detected genes log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
plotColData(sce, x = "Sample", y = "sum") +
  scale_y_log10() + ggtitle("Total count log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
```

Split by cell type

```{r violincluster, eval=FALSE}
#mt
plotColData(sce, x = "cluster_names", y = "subsets_mt_percent", colour_by = "Sample") +
scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
   ggtitle("mt genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)
#detected
plotColData(sce, x = "cluster_names", y = "detected",  colour_by = "Sample") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
   ggtitle("Detected genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)
#sum
plotColData(sce, x = "cluster_names", y = "sum",  colour_by = "Sample") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  ggtitle("Total count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)
```

Split by cell type, log scales

```{r violinglog, eval=FALSE}
plotColData(sce, x = "cluster_names", y = "detected",  colour_by = "Sample") +
  scale_y_log10(breaks = scales::pretty_breaks(n = 12)) + 
  ggtitle("Detected genes log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols) 


plotColData(sce, x = "cluster_names", y = "sum",  colour_by = "Sample") +
 scale_y_log10(breaks = scales::pretty_breaks(n = 12))+
  ggtitle("Total count log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = cols)
```

Split by sample type, facet wrap

```{r violinwrap, fig.height=14, fig.width=14, eval=TRUE}

# detected
plotColData(sce, x = "Sample", y = "detected",  colour_by = "genotype", other_fields = "cluster_names") +
  scale_y_log10(breaks = scales::pretty_breaks(n = 12)) + 
  ggtitle("Detected genes log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +           scale_color_manual(values = col_wt_het_ko)  + facet_wrap(~cluster_names)
# sum
plotColData(sce, x = "Sample", y = "sum",  colour_by = "genotype", other_fields = "cluster_names") +
 scale_y_log10(breaks = scales::pretty_breaks(n = 12))+
  ggtitle("Total count log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = col_wt_het_ko)+ facet_wrap(~cluster_names)

```

```{r hist}
hist(
  sce$detected,
  breaks = 100
)
hist(
  sce$sum,
  breaks = 100
)
hist(
  sce$subsets_mt_percent,
  breaks = 100
)
```

### Scatter plots

```{r scatter, fig.height=14, fig.width=14, eval=FALSE}

plotColData(sce, x = "sum", y = "subsets_mt_percent", colour_by = "genotype", other_fields = "cluster_names") + facet_wrap(~cluster_names) + 
  scale_color_manual(values = col_wt_het_ko)

plotColData(sce, x = "sum", y = "detected", colour_by = "genotype", other_fields = "cluster_names") + facet_wrap(~cluster_names) + 
  scale_color_manual(values = col_wt_het_ko)

```

```{r scatter2, fig.height=14, fig.width=14, eval=TRUE}

plotColData(sce, x = "sum", y = "subsets_mt_percent", colour_by = "Sample", other_fields = "cluster_names") + facet_wrap(~cluster_names) + 
  scale_color_manual(values = cols)

plotColData(sce, x = "sum", y = "detected", colour_by = "Sample", other_fields = "cluster_names") + facet_wrap(~cluster_names) + 
  scale_color_manual(values = cols)

```

</details>

We set the thresholds for each celltype:

**Sum- Detected**

OPCs 3000 -1500

oligo 5000-1500

Astrocyte 2000 -1000

Microglia 3000 -1500

endothelial keep same

immature_neurons 4000 -1500

mature_neurons 2500 -1500

BAMS keep same

oligo_astrocytes keep same

immune 2000 -1000

pericytes 2000 -1000

ChP_epithelial 2000 -1000

**Mt**

all 10% except astrocytes, keep the same

```{r subset, eval=TRUE}

# create vector with same levels as the celltype to replace by thresholds for each
sce$celltype <- droplevels(sce$celltype)
sce$threshold_umi <- (sce$celltype)
sce$threshold_detected <- (sce$celltype)
sce$threshold_mt <- sce$celltype
#to get the order:
#dput(levels(sce$threshold_umi))
#c("OPC", "oligo", "astrocyte", "microglia", "OPC?", "endothelial", "immature_neurons", "mature_neurons", "mono", "pericytes", "BAMs", "ChP_epithelial", "oligo-astros")

# replace 
levels(sce$threshold_umi) <- c(3000, 5000, 2000, 3000, 3000, 462, 
4000, 2500, 2000, 2000, 462, 
2000, 462)
levels(sce$threshold_detected) <- c(1500, 1500, 1000, 1500, 1500, 429, 
1500, 1500, 1000, 1000, 462, 
1000, 462)
levels(sce$threshold_mt) <- c(10, 10, 15, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10)

# transform the factor into numeric
sce$threshold_detected <- as.numeric(as.character(sce$threshold_detected))
sce$threshold_umi <- as.numeric(as.character(sce$threshold_umi))
sce$threshold_mt <- as.numeric(as.character(sce$threshold_mt))

# set thresholds
discard_umi <- sce$sum < sce$threshold_umi
discard_detected <- sce$detected < sce$detected
discard_mt <- sce$subsets_mt_percent > sce$threshold_mt
discard <- discard_umi | discard_detected | discard_mt

# subset
sce <- sce[,!discard]
```

We discard `r sum(discard_umi)` cells due to umi counts,
`r sum(discard_detected)` due to detected genes and `r sum(discard_mt)`
cells for high mt percentatge.

in total we discard `r sum(discard)` cells

# Gene QC

In the preliminary analysis we only deleted genes that had less than 2
cells expressing it, now we filter all genes with less than 10 cells
expressing it

```{r gene, eval=TRUE}
# at least 10 cells should express the gene
keep_feature <- rowSums(counts(sce) > 0) > 10
sce <- sce[keep_feature,]
```

We keep `r sum(keep_feature)` genes

```{r}
if(!file.exists(here("processed", project,  "sce_QC_02.RDS"))){
  saveRDS(sce, here("processed", project,  "sce_QC_02.RDS"))
  }
```

Final dimension is  `r dim(sce)`

# Normalisation

```{r normalisation}
## Normalise by deconvolution ---
if(!file.exists(here("processed", project,  "sce_norm_02.RDS"))){
# For reproducibility
set.seed(100)
# Quick clustering to pool samples together and deal with 0 counts
quick_clusters <- quickCluster(sce)
# Calculate size factors
sce <-
  computeSumFactors(sce, cluster = quick_clusters, min.mean = 0.1)
# Check that there are not negative size factors
summary(sizeFactors(sce))
# Apply size factors and log transform them
sce <- logNormCounts(sce)
# save object
saveRDS(sce, here("processed", project,  "sce_norm_02.RDS"))
}
```

# Feature selection

After the quality control the variable genes need to be selected again
to better represent the variance in this cleaned dataset. We follow the
same methods than for our first feature selection, selecting here for
the top 2000 genes.

```{r featureselect}
gene_var_df <- modelGeneVar(sce)
gene_var <- metadata(gene_var_df)
plot(gene_var$mean, gene_var$var, xlab= "Mean of log-expression", ylab= "Variance of log-expression")
curve(gene_var$trend(x), lwd=2, add=T, col = "red")
# select hvgs
hvgs <- getTopHVGs(gene_var_df, n=2000)
# save them in the object
rowSubset(sce) <- hvgs
```

# Dimensionality reduction

## Run PCA and choose PCs

```{r }
if(!file.exists(here("processed", project,  "sce_dimred_02.RDS"))){
set.seed(1000)
sce <- runPCA(sce)

pct_var <- attr(reducedDim(sce, "PCA"), "percentVar")
plot(pct_var, log="y", xlab="PC", ylab="pct variance explained")



 #will keep 28PCs 

#keep the previous dimensional reduction just in case
reducedDim(sce, "PCA_all") <- reducedDim(sce, "PCA")
# And replace the default PCA with the reduced PCs
reducedDim(sce, "PCA") <- reducedDim(sce, "PCA")[,1:28]
}
```

## Visualisation

For visualisation, reduce to 2 dimensions. Non linear reductions.

### UMAP and TSNE

```{r eval=TRUE}
if(!file.exists(here("processed", project,  "sce_dimred_02.RDS"))){
set.seed(1000)
sce <- runUMAP(sce, dimred="PCA")
plotReducedDim(sce, dimred="UMAP", colour_by="genotype", point_size=0.1, point_alpha = 0.3)
set.seed(1000)
sce <- runTSNE(sce, dimred="PCA")
plotReducedDim(sce, dimred="TSNE", colour_by="genotype", point_size=0.1, point_alpha = 0.3)
plotReducedDim(sce, dimred="TSNE", colour_by="chip", point_size=0.1, point_alpha = 0.3)


saveRDS(sce, here("processed", project, "sce_dimred_02.RDS"))

}else{ # load teh corrected object
  sce <- readRDS(here("processed", project,  "sce_corrected_02.RDS"))
  plotReducedDim(sce, "TSNE_uncorrected", colour_by="chip")
}
```

## Rerrun batch correction

```{r batch, eval=TRUE, message=FALSE, warning=FALSE}
if (!(file.exists(
  here("processed", project,  "sce_corrected_02.RDS")
))) {
set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$chip),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  merge.order = 
    list(list("3","5"), list("4","6")),
  d = 23,
  prop.k=0.10
  )
)
# recompute dimensional reduction 
#keeping  the previous dimensional reduction
reducedDim(sce, "TSNE_uncorrected") <- reducedDim(sce, "TSNE")
reducedDim(sce, "UMAP_uncorrected") <- reducedDim(sce, "UMAP")
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")

set.seed(100)
sce <- runUMAP(sce, dimred="corrected")

# save
saveRDS( sce, here("processed", project,  "sce_corrected_02.RDS"))
}
plotReducedDim(sce, colour_by= "chip", dimred = "TSNE") + 
      ggtitle("TSNE dimensional reduction corrected")
plotReducedDim(sce, dimred="TSNE", colour_by="celltype")
```
