---
title: "Trajectory Oligos"
author: "Nadine Bestard"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set-up
```{r}
library(here) # paths
library(scran) # feature select
library(batchelor) # correct batch effect
library(scater) # dimred calculation
library(bluster) # clustering
library(slingshot) # trajectory
```

```{r}
project <- "fire-mice"
source(here("src/colours.R"))
```

## Subset


```{r subset}
sce <- readRDS(here("processed", project, "sce_anno_02.RDS"))
oligodendroglia <- c("OPC", "pOPC", "iOligo", "mOligo_1", "mOligo_2", "mOligo_3", 
"mOligo_4")
sce <- sce[,sce$clusters_named %in% oligodendroglia]

# remove unnecessary slots that are not representative anymore
for(dim in reducedDimNames(sce)){
  reducedDim(sce, dim) <- NULL
}
clusters_k <- grep("cluster_k", colnames(colData(sce)), value = T)
clusters_seurat <- grep("originalexp_snn_res",colnames(colData(sce)), value = T)
for(meta in c(clusters_seurat, clusters_k, "cluster_names")){
  colData(sce)[[meta]] <- NULL
}
assay(sce, "reconstructed") <- NULL

```

As the QC thresholds in "QC_norm_featureselect_dimred_02.Rmd" are already celltype specific,
ther is no need to redo a cell QC after subsetting. however some genes might not be expressed in this subsetted dataset, and will be removed for downstream. same threshold as in the past is used, mimimum 10 cells have to express the gene in order to keep it. 

```{r genes}
keep_feature <- rowSums(counts(sce) > 0) > 10
sce <- sce[keep_feature,]
```

# Feature selection

After subsetting the data the variable genes need to be selected again
to better represent the variance in this cell population. We follow the
same methods than for our first feature selection, selecting here for
the top 2000 genes.

```{r featureselect}
gene_var_df <- modelGeneVar(sce, density.weights = FALSE) # density weights avoids overfitting due to high-abundance genes being HV
gene_var <- metadata(gene_var_df)
plot(gene_var$mean, gene_var$var, xlab= "Mean of log-expression", ylab= "Variance of log-expression")
curve(gene_var$trend(x), lwd=2, add=T, col = "red")
# select hvgs
hvgs <- getTopHVGs(gene_var_df, n=2000)
# save them in the object
rowSubset(sce) <- hvgs
```
# Batch correction

The Dimensional reduction step is skipped as we want to work on the corrected space. 

```{r}
set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$chip),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  merge.order = 
    list(list("3","5"), list("4","6")),
  d = 23,
  prop.k=0.10
  )
)
# compute dimensional reduction 
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")

set.seed(100)
sce <- runUMAP(sce, dimred="corrected")

plotReducedDim(sce, "UMAP", colour_by = "chip")
plotReducedDim(sce, "UMAP", colour_by = "clusters_named")
plotReducedDim(sce, "TSNE", colour_by = "clusters_named")
```


# Trajectory analysis with SlingShot


```{r}
# https://github.com/kstreet13/slingshot/issues/87
reducedDim(sce, "corrected-1") <- reducedDim(sce, "corrected")[,1:22]
# run slingshot
sce <- slingshot(sce, clusterLabels = "clusters_named", reducedDim= "corrected-1", start.clus="OPC")
summary(sce$slingPseudotime_1)
```


```{r}
# as there are 3 trajectory lineages a mean between them can be calucluated
sce$slingPseudotimes <- rowMeans(cbind(sce$slingPseudotime_1, sce$slingPseudotime_2, sce$slingPseudotime_3), na.rm = TRUE)

plotReducedDim(sce, "UMAP", colour_by = "slingPseudotimes") 

# plot the lineages
cols <- pals::viridis(100)[cut(sce$slingPseudotimes, breaks = 100)]
plot(reducedDims(sce)$corrected, col = cols, ylab = "PCA 2", xlab = "PCA 1")
lines(SlingshotDataSet(sce), lwd=2, col='black', type = "lineages")

# plot the smooth lines
cols <- pals::viridis(100)[cut(sce$slingPseudotimes, breaks = 100)]
plot(reducedDims(sce)$corrected, col = cols, ylab = "PCA 2", xlab = "PCA 1")
lines(SlingshotDataSet(sce), lwd=2, col='black')
# map back to what they correspond
plotReducedDim(sce, "corrected", colour_by= "clusters_named")  + ylab ("PCA 2") + xlab("PCA 1")

# lineages
slingLineages(SlingshotDataSet(sce))
```
Two lineages are found (+ going to the pOPCs). They go through oligo4. 

Map the trajectory back to the UMAP. this is the "smooth curve"option. Solution from Slingshot pdf. 
It looks better on the other embedding
```{r}
po <- embedCurves(sce, "UMAP")
plot(reducedDims(sce)$UMAP, col = cols, asp = 1)
lines(SlingshotDataSet(po), lwd = 3)
```


```{r, eval=F}
# extract information from slingshot object
embed <- embedCurves(sce, "UMAP")
embed <- slingCurves(embed)[[1]] 
embed <- data.frame(embed$s[embed$ord, ])

# i need to figure out how to plot the lineages
plotUMAP(sce, colour_by = "slingPseudotime_1") +
  geom_path(data = embed, aes(x = Dim.1, y = Dim.2), size = 1.2)

# lineages
slingLineages(SlingshotDataSet(sce))
# specific cluster for KO is marked as a transitory stage.
```

```{r}
# This can be modified to plot plus or less lines! ###
# loop over the paths and add each one separately.
gg <- plotReducedDim(sce, dimred = "corrected", colour_by="slingPseudotimes")
#embedded <- embedCurves(sce, "corrected")
embedded <- SlingshotDataSet(sce)
embedded <- slingCurves(embedded)
for (path in embedded) {
    embedded_l <- data.frame(path$s[path$ord,])
    gg <- gg + geom_path(data=embedded_l, aes(x=Dim.1, y=Dim.2), size=1.2)
}

gg
```

```{r eval=FALSE, include=FALSE}
#NOT WORKING
# create a new psuedo time,mean of the two paths
sce$slingPseudotimes <- rowMeans(cbind(sce$slingPseudotime_1, sce$slingPseudotime_2))



# extract information from slingshot object
embed <- embedCurves(sce, "UMAP")
embed <- slingCurves(embed)

# plot umap
umap <- plotUMAP(sce, colour_by = "slingPseudotime_1")
# plot each lineage
for (path in embed) {
embed_l <- data.frame(path$s[path$ord, ])

# i need to figure out how to plot the 
 plot <- umap +
  geom_path(data = embed_l, aes(x = Dim.1, y = Dim.2), size = 1.2)
}

plot
# lineages
slingLineages(SlingshotDataSet(sce))
```

```{r, eval=F}
# From OSCA
sce$slingPseudotimes <- rowMeans(cbind(sce$slingPseudotime_1, sce$slingPseudotime_2), na.rm = TRUE)
# Need to loop over the paths and add each one separately.
gg <- plotUMAP(sce, colour_by="slingPseudotimes")
embedded <- embedCurves(sce, "UMAP")
embedded <- slingCurves(embedded)
for (path in embedded) {
    embedded_l <- data.frame(path$s[path$ord,])
    gg <- gg + geom_path(data=embedded_l, aes(x=Dim.1, y=Dim.2), size=1.2)
}

gg
```
Ok, the plottings are terrible, but the information is there. 

# Slingshot on the UMAP space.

This is the way to go to get nice embedded lines in the UMAP plot. 

```{r}
sce_umap <- sce
previous <- grep("slingPseudotime", colnames(colData(sce_umap)), value=T)
for(meta in previous){
  colData(sce_umap)[[meta]] <- NULL
}
# run slingshot
sce_umap <- slingshot(sce, clusterLabels = "clusters_named", reducedDim= "UMAP", start.clus="OPC")
summary(sce$slingPseudotime_1)
plotReducedDim(sce_umap, "UMAP", colour_by = "slingPseudotime_1") 


# create a new psuedo time,mean of the two paths
sce_umap$slingPseudotimes <- rowMeans(cbind(sce_umap$slingPseudotime_1, sce_umap$slingPseudotime_2, sce_umap$slingPseudotime_3), na.rm = TRUE)
#sce_umap$slingPseudotimes[is.na(sce_umap$slingPseudotimes)] <- min(sce$slingPseudotimes)
cols <- pals::viridis(100)[cut(sce_umap$slingPseudotimes, breaks = 100)]
plot(reducedDims(sce_umap)$UMAP, col = cols)
lines(SlingshotDataSet(sce_umap), lwd=2, col='black', type = "lineages")


#plotReducedDim(sce, "corrected", colour_by= "clusters_named")
```

Just... no. No. iOligos should be in the path. 

I tried another seed for the umap, where the oligo cluster is upside-down, and then the starting cluster is suddenly oligo-3 and then it branches to oligo4, oligo2 and oligo1. 

- Conclusion: No Umap .Backed up by [this](https://www.youtube.com/watch?v=Fbd08bIv4yk&t=776s), from the developers.

```{r save}
saveRDS(sce, here("processed", project, "oligo_trajectory.rds"))
```

# Plotting density

```{r}
# Pseudotime densities (by treatment)
ds <- list(FIRE = density(sce$slingPseudotimes[colData(sce)$genotype == "KO"]),
           WT = density(sce$slingPseudotimes[colData(sce)$genotype == "WT"]),
           HET = density(sce$slingPseudotimes[colData(sce)$genotype == "HET"]) )
xlim <- range(c(ds$WT$x, ds$FIRE$x, ds$HET$x))
ylim <- range(c(ds$WT$y, ds$FIRE$y, ds$HET$y))
plot(xlim, ylim, col = "white", xlab = "Pseudotime", ylab = "")
polygon(c(min(ds$WT$x),ds$WT$x,max(ds$WT$x)),
		c(0,ds$WT$y,0), col = alpha(col_wt_het_ko[1], alpha = .5))
polygon(c(min(ds$HET$x),ds$HET$x,max(ds$HET$x)),
		c(0,ds$HET$y,0), col = alpha(col_wt_het_ko[2], alpha = .5))
polygon(c(min(ds$FIRE$x),ds$FIRE$x,max(ds$FIRE$x)),
		c(0,ds$FIRE$y,0), col = alpha(col_wt_het_ko[3], alpha = .5))
legend("topright", legend = c("WT","HET", "FIRE"), 
       fill = alpha(c(col_wt_het_ko[1], col_wt_het_ko[2],col_wt_het_ko[3]), alpha = .5), bty = "n")
```
```{r}
# separate
plot(xlim, ylim, col = "white", xlab = "Pseudotime", ylab = "density WT")
polygon(c(min(ds$WT$x),ds$WT$x,max(ds$WT$x)),
		c(0,ds$WT$y,0), col = alpha(col_wt_het_ko[1], alpha = .5))
#legend("topright", legend = c("WT","HET", "FIRE"), 
#       fill = alpha(c(col_wt_het_ko[1], col_wt_het_ko[2],col_wt_het_ko[3]), alpha = .5), bty = "n")


plot(xlim, ylim, col = "white", xlab = "Pseudotime", ylab = "density HET")
polygon(c(min(ds$HET$x),ds$HET$x,max(ds$HET$x)),
		c(0,ds$HET$y,0), col = alpha(col_wt_het_ko[2], alpha = .5))

plot(xlim, ylim, col = "white", xlab = "Pseudotime", ylab = "density FIRE")
polygon(c(min(ds$FIRE$x),ds$FIRE$x,max(ds$FIRE$x)),
		c(0,ds$FIRE$y,0), col = alpha(col_wt_het_ko[3], alpha = .5))
```
```{r}
# test for significance Kolmogorov-Smirnov Test (https://kstreet13.github.io/bioc2020trajectories/articles/workshopTrajectories.html)
ks.test(sce$slingPseudotimes[colData(sce)$genotype == "WT"]*100,
        sce$slingPseudotimes[colData(sce)$genotype == "KO"]*100)

ks.test(sce$slingPseudotimes[colData(sce)$genotype == "WT"],
        sce$slingPseudotimes[colData(sce)$genotype == "HET"])


```
More information about KS test:
<details>
https://en.wikipedia.org/wiki/Kolmogorov%E2%80%93Smirnov_test
https://stats.stackexchange.com/questions/333892/is-the-kolmogorov-smirnov-test-too-strict-if-the-sample-size-is-large
https://stats.stackexchange.com/questions/7771/how-do-i-calculate-the-effect-size-for-the-kolmogorov-smirnov-z-statistic

In this test, the statistic (D) indeed also reflects the effect size. The interpretation is that D is the maximum distance between the two cumulative distributions from our data. A big number, reflects a big distance, big effect; small number small effect. The maximum value is 1. 
```{r}
plot(ecdf(sce$slingPseudotimes[colData(sce)$genotype == "WT"]))
lines(ecdf(sce$slingPseudotimes[colData(sce)$genotype == "KO"]), col="red")

```

This is how the cumulative distributions look for the FIRE (red) and WT (black). The  maximum distance is arround pseudotime (x) 1.5.

"there is a difference, but not sure it is biologically relevant. My first impression as well as yours was "these curves look similar". The test will be significant with the tiniest of differences as the sample number is so big, as you well pointed and is also mentioned in the link I forwarded about picking noise.
So, it is up to our human expertise to decide if the effect size is biologically relevant or not. Then, why not going back to our first human impression which was "these curves look similar"? "
For other possible interpretations go back to email thread.

</details>

# Separate WT and KO

## WT

```{r subset-preprocess, eval=F}
## this would be interesting, but actually it's better to keep the same graph, so we can compare more easily. 
sce_WT <- sce[, sce$genotype == "WT"]
# remove the previous slingshot run
previous <- grep("slingPseudotime", colnames(colData(sce_WT)), value=T)
for(meta in previous){
  colData(sce_WT)[[meta]] <- NULL
}
keep_feature <- rowSums(counts(sce_WT) > 0) > 10
sce_WT <- sce_WT[keep_feature,]
gene_var_df <- modelGeneVar(sce_WT, density.weights = FALSE) # density weights avoids overfitting due to high-abundance genes being HV
gene_var <- metadata(gene_var_df)
plot(gene_var$mean, gene_var$var, xlab= "Mean of log-expression", ylab= "Variance of log-expression")
curve(gene_var$trend(x), lwd=2, add=T, col = "red")
# select hvgs
hvgs <- getTopHVGs(gene_var_df, n=2000)
# save them in the object
rowSubset(sce_WT) <- hvgs


#recompute the corrected graph 

set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$chip),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  merge.order = 
    list(list("3","5"), list("4","6")),
  d = 23,
  prop.k=0.10
  )
)
# compute dimensional reduction 
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")

set.seed(100)
sce <- runUMAP(sce, dimred="corrected")

plotReducedDim(sce, "UMAP", colour_by = "chip")
plotReducedDim(sce, "UMAP", colour_by = "clusters_named")
# https://github.com/kstreet13/slingshot/issues/87
reducedDim(sce, "corrected-1") <- reducedDim(sce, "corrected")[,1:22]
```

```{r subset-wt, eval=T}
sce_WT <- sce[, sce$genotype == "WT"]
# remove the previous slingshot run
previous <- grep("slingPseudotime", colnames(colData(sce_WT)), value=T)
for(meta in previous){
  colData(sce_WT)[[meta]] <- NULL
}
```


```{r}
# run slingshot
sce_WT <- slingshot(sce_WT, clusterLabels = "clusters_named", reducedDim= "corrected-1", start.clus="OPC")
summary(sce_WT$slingPseudotime_1)

```


```{r}
# as there are 3 trajectory lineages a mean between them can be calucluated
sce_WT$slingPseudotimes <- rowMeans(cbind(sce_WT$slingPseudotime_1, sce_WT$slingPseudotime_2, sce_WT$slingPseudotime_3), na.rm = TRUE)

plotReducedDim(sce_WT, "UMAP", colour_by = "slingPseudotimes") 

# plot the lineages
cols <- pals::viridis(100)[cut(sce_WT$slingPseudotimes, breaks = 100)]
plot(reducedDims(sce_WT)$corrected, col = cols)
lines(SlingshotDataSet(sce_WT), lwd=2, col='black', type = "lineages")

# plot the smooth lines
cols <- pals::viridis(100)[cut(sce_WT$slingPseudotimes, breaks = 100)]
plot(reducedDims(sce_WT)$corrected, col = cols)
lines(SlingshotDataSet(sce_WT), lwd=2, col='black')
# map back to what they correspond
plotReducedDim(sce_WT, "corrected", colour_by= "clusters_named") + ylab ("PCA 2") + xlab("PCA 1")

# lineages
slingLineages(SlingshotDataSet(sce_WT))
```

## KO
```{r subset-ko, eval=T}
sce_KO <- sce[, sce$genotype == "KO"]
# remove the previous slingshot run
previous <- grep("slingPseudotime", colnames(colData(sce_KO)), value=T)
for(meta in previous){
  colData(sce_KO)[[meta]] <- NULL
}
```


```{r}
# run slingshot
sce_KO <- slingshot(sce_KO, clusterLabels = "clusters_named", reducedDim= "corrected-1", start.clus="OPC")
summary(sce_KO$slingPseudotime_1)

```


```{r}
# as there are 2 trajectory lineages a mean between them can be calucluated
sce_KO$slingPseudotimes <- rowMeans(cbind(sce_KO$slingPseudotime_1, sce_KO$slingPseudotime_2), na.rm = TRUE)

plotReducedDim(sce_KO, "UMAP", colour_by = "slingPseudotimes") 

# plot the lineages
cols <- pals::viridis(100)[cut(sce_KO$slingPseudotimes, breaks = 100)]
plot(reducedDims(sce_KO)$corrected, col = cols, ylab = "PCA 2", xlab = "PCA 1")
lines(SlingshotDataSet(sce_KO), lwd=2, col='black', type = "lineages")

# plot the smooth lines
cols <- pals::viridis(100)[cut(sce_KO$slingPseudotimes, breaks = 100)]
plot(reducedDims(sce_KO)$corrected, col = cols, ylab = "PCA 2", xlab = "PCA 1")
lines(SlingshotDataSet(sce_KO), lwd=2, col='black')
# map back to what they correspond
plotReducedDim(sce_KO, "corrected", colour_by= "clusters_named") + ylab ("PCA 2") + xlab("PCA 1")

# lineages
slingLineages(SlingshotDataSet(sce_KO)) 
```

<details>

<summary> Click to expand </summary>

```{r session-info}
sessionInfo()
```

</details>