---
title: "Trajectory Oligos"
author: "Nadine Bestard"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set-up
```{r}
library(here) # paths
library(scran) # feature select
library(batchelor) # correct batch effect
library(scater) # dimred calculation
library(bluster) # clustering
library(slingshot) # trajectory
```

```{r}
project <- "fire-mice"
source(here("src/colours.R"))
```

## Subset


```{r subset}
sce <- readRDS(here("processed", project, "sce_anno_02.RDS"))
oligodendroglia <- c("OPC", "pOPC", "iOligo", "mOligo_1", "mOligo_2", "mOligo_3", 
"mOligo_4")
sce <- sce[,sce$clusters_named %in% oligodendroglia]

# remove unnecessary slots that are not representative anymore
for(dim in reducedDimNames(sce)){
  reducedDim(sce, dim) <- NULL
}
clusters_k <- grep("cluster_k", colnames(colData(sce)), value = T)
clusters_seurat <- grep("originalexp_snn_res",colnames(colData(sce)), value = T)
for(meta in c(clusters_seurat, clusters_k, "cluster_names")){
  colData(sce)[[meta]] <- NULL
}
assay(sce, "reconstructed") <- NULL

```

As the QC thresholds in "QC_norm_featureselect_dimred_02.Rmd" are already celltype specific,
ther is no need to redo a cell QC after subsetting. however some genes might not be expressed in this subsetted dataset, and will be removed for downstream. same threshold as in the past is used, mimimum 10 cells have to express the gene in order to keep it. 

```{r genes}
keep_feature <- rowSums(counts(sce) > 0) > 10
sce <- sce[keep_feature,]
```

# Feature selection

After subsetting the data the variable genes need to be selected again
to better represent the variance in this cell population. We follow the
same methods than for our first feature selection, selecting here for
the top 2000 genes.

```{r featureselect}
gene_var_df <- modelGeneVar(sce, density.weights = FALSE) # density weights avoids overfitting due to high-abundance genes being HV
gene_var <- metadata(gene_var_df)
plot(gene_var$mean, gene_var$var, xlab= "Mean of log-expression", ylab= "Variance of log-expression")
curve(gene_var$trend(x), lwd=2, add=T, col = "red")
# select hvgs
hvgs <- getTopHVGs(gene_var_df, n=2000)
# save them in the object
rowSubset(sce) <- hvgs
```
# Batch correction

The Dimensional reduction step is skipped as we want to work on the corrected space. 

```{r}
set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$chip),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  merge.order = 
    list(list("3","5"), list("4","6")),
  d = 23,
  prop.k=0.10
  )
)
# compute dimensional reduction 
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")

set.seed(100)
sce <- runUMAP(sce, dimred="corrected")

plotReducedDim(sce, "UMAP", colour_by = "chip")
plotReducedDim(sce, "UMAP", colour_by = "clusters_named")
```

# clustering
```{r}
ks <- c(5, 10, 15, 20, 25, 30, 35, 40)
clust <- clusterSweep(reducedDim(sce, "corrected"),
                      NNGraphParam(),
                      k = as.integer(c(5, 10, 15, 20, 25, 30, 35, 40)))

names <- paste0("k.", ks)
# Use the list to store the info in the sce and plot result
  idx <- c(1:length(names))
  for (i in idx) {
    k <- ks[i]
    cluster_name <- names[i]
    # remove previous versions of this clustering 
    colData(sce)[[cluster_name]] <- NULL
    colData(sce) <- cbind(colData(sce), clust$clusters[cluster_name])
    # and plot the result
    plotDim <-
      plotReducedDim(
        sce,
        "UMAP",
        colour_by = cluster_name,
        point_size = 0.5,
        point_alpha = 0.3,
        text_by = cluster_name,
        text_size = 3
      )  + scale_color_hue() + ggtitle(paste("Clustering with k =", k))
    print(plotDim)
  }
  
  saveRDS(here("processed", project, "oligo_clusters"))
```

# Trajectory analysis with SlingShot


```{r}
# https://github.com/kstreet13/slingshot/issues/87
reducedDim(sce, "corrected-1") <- reducedDim(sce, "corrected")[,1:22]
# run slingshot
sce <- slingshot(sce, clusterLabels = "k.40", reducedDim= "corrected-1")
summary(sce$slingPseudotime_1)
plotReducedDim(sce, "UMAP", colour_by = "slingPseudotime_1") 

cols <- pals::viridis(100)[cut(sce$slingPseudotime_1, breaks = 100)]

plot(reducedDims(sce)$corrected, col = cols)
lines(SlingshotDataSet(sce), lwd=2, col='black', type = "lineages")

plotReducedDim(sce, "corrected", colour_by= "clusters_named")
```
Only one lineage was found. 

Map the trajectory back to the UMAP. this is the "smooth curve"option. Still don't know how to plot the lineages

```{r}
# extract information from slingshot object
embed <- embedCurves(sce, "UMAP")
embed <- slingCurves(embed)[[1]] 
embed <- data.frame(embed$s[embed$ord, ])

# i need to figure out how to plot the lineages
plotUMAP(sce, colour_by = "slingPseudotime_1") +
  geom_path(data = embed, aes(x = Dim.1, y = Dim.2), size = 1.2)

# lineages
slingLineages(SlingshotDataSet(sce))
# specific cluster for KO is marked as a transitory stage.
```
TODO: 
improve plotting.
mark fire mice as end point, then it could : branch or be at the end of this single lineage - DONE

Separate WT and KO ( cluster specific for KO will not be in the WT, how this affects trajectory? )


Marking fire-mice specific cluster as end-point, as we know that these cells are present in the tissue and 
aren't just a transitionary state.
```{r}
# run slingshot with end point
sce <- slingshot(sce, clusterLabels = "k.40", reducedDim= "corrected-1", end.clus = "9")
summary(sce$slingPseudotime_1)
plotReducedDim(sce, "UMAP", colour_by = "slingPseudotime_1") 

cols <- pals::viridis(100)[cut(sce$slingPseudotime_1, breaks = 100)]

plot(reducedDims(sce)$corrected, col = cols)
lines(SlingshotDataSet(sce), lwd=2, col='black', type = "lineages")

# lineages
slingLineages(SlingshotDataSet(sce))
# Now the KO specific state forced the creation of a new lineage. 
```

```{r eval=FALSE, include=FALSE}
#NOT WORKING
# create a new psuedo time,mean of the two paths
sce$slingPseudotimes <- rowMeans(cbind(sce$slingPseudotime_1, sce$slingPseudotime_2))



# extract information from slingshot object
embed <- embedCurves(sce, "UMAP")
embed <- slingCurves(embed)

# plot umap
umap <- plotUMAP(sce, colour_by = "slingPseudotime_1")
# plot each lineage
for (path in embed) {
embed_l <- data.frame(path$s[path$ord, ])

# i need to figure out how to plot the 
 plot <- umap +
  geom_path(data = embed_l, aes(x = Dim.1, y = Dim.2), size = 1.2)
}

plot
# lineages
slingLineages(SlingshotDataSet(sce))
```

```{r}
# From OSCA
# Need to loop over the paths and add each one separately.
gg <- plotUMAP(sce, colour_by="slingPseudotimes")
embedded <- embedCurves(sce, "UMAP")
embedded <- slingCurves(embedded)
for (path in embedded) {
    embedded_l <- data.frame(path$s[path$ord,])
    gg <- gg + geom_path(data=embedded_l, aes(x=Dim.1, y=Dim.2), size=1.2)
}

gg
```
Ok, the plottings are terrible, but the information is there. 

