---
title: "Trajectory Oligos"
author: "Nadine Bestard"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set-up
```{r}
library(here) # paths
library(scran) # feature select
library(batchelor) # correct batch effect
library(scater) # dimred calculation
library(bluster) # clustering
library(slingshot) # trajectory
```

```{r}
project <- "fire-mice"
source(here("src/colours.R"))
```

## Subset


```{r subset}
sce <- readRDS(here("processed", project, "sce_anno_02.RDS"))
oligodendroglia <- c("OPC", "pOPC", "iOligo", "mOligo_1", "mOligo_2", "mOligo_3", 
"mOligo_4")
sce <- sce[,sce$clusters_named %in% oligodendroglia]

# remove unnecessary slots that are not representative anymore
for(dim in reducedDimNames(sce)){
  reducedDim(sce, dim) <- NULL
}
clusters_k <- grep("cluster_k", colnames(colData(sce)), value = T)
clusters_seurat <- grep("originalexp_snn_res",colnames(colData(sce)), value = T)
for(meta in c(clusters_seurat, clusters_k, "cluster_names")){
  colData(sce)[[meta]] <- NULL
}
assay(sce, "reconstructed") <- NULL

```

As the QC thresholds in "QC_norm_featureselect_dimred_02.Rmd" are already celltype specific,
ther is no need to redo a cell QC after subsetting. however some genes might not be expressed in this subsetted dataset, and will be removed for downstream. same threshold as in the past is used, mimimum 10 cells have to express the gene in order to keep it. 

```{r genes}
keep_feature <- rowSums(counts(sce) > 0) > 10
sce <- sce[keep_feature,]
```

# Feature selection

After subsetting the data the variable genes need to be selected again
to better represent the variance in this cell population. We follow the
same methods than for our first feature selection, selecting here for
the top 2000 genes.

```{r featureselect}
gene_var_df <- modelGeneVar(sce, density.weights = FALSE) # density weights avoids overfitting due to high-abundance genes being HV
gene_var <- metadata(gene_var_df)
plot(gene_var$mean, gene_var$var, xlab= "Mean of log-expression", ylab= "Variance of log-expression")
curve(gene_var$trend(x), lwd=2, add=T, col = "red")
# select hvgs
hvgs <- getTopHVGs(gene_var_df, n=2000)
# save them in the object
rowSubset(sce) <- hvgs
```
# Batch correction

The Dimensional reduction step is skipped as we want to work on the corrected space. 

```{r}
set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$chip),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  merge.order = 
    list(list("3","5"), list("4","6")),
  d = 23,
  prop.k=0.10
  )
)
# compute dimensional reduction 
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")

set.seed(100)
sce <- runUMAP(sce, dimred="corrected")

plotReducedDim(sce, "UMAP", colour_by = "chip")
plotReducedDim(sce, "UMAP", colour_by = "clusters_named")
plotReducedDim(sce, "TSNE", colour_by = "clusters_named")
```


# Trajectory analysis with SlingShot


```{r}
# https://github.com/kstreet13/slingshot/issues/87
reducedDim(sce, "corrected-1") <- reducedDim(sce, "corrected")[,1:22]
# run slingshot
sce <- slingshot(sce, clusterLabels = "clusters_named", reducedDim= "corrected-1", start.clus="OPC")
summary(sce$slingPseudotime_1)
plotReducedDim(sce, "UMAP", colour_by = "slingPseudotime_1") 

cols <- pals::viridis(100)[cut(sce$slingPseudotime_1, breaks = 100)]

plot(reducedDims(sce)$corrected, col = cols)
lines(SlingshotDataSet(sce), lwd=2, col='black', type = "lineages")

plotReducedDim(sce, "UMAP", colour_by= "clusters_named")

# lineages
slingLineages(SlingshotDataSet(sce))
```
Two lineages are found. They go through oligo4. 

Map the trajectory back to the UMAP. this is the "smooth curve"option. Solution from Slingshot pdf. 
```{r}
po <- embedCurves(sce, "UMAP")
plot(reducedDims(sce)$UMAP, col = cols, asp = 1)
lines(SlingshotDataSet(po), lwd = 3)
```


```{r}
# extract information from slingshot object
embed <- embedCurves(sce, "UMAP")
embed <- slingCurves(embed)[[1]] 
embed <- data.frame(embed$s[embed$ord, ])

# i need to figure out how to plot the lineages
plotUMAP(sce, colour_by = "slingPseudotime_1") +
  geom_path(data = embed, aes(x = Dim.1, y = Dim.2), size = 1.2)

# lineages
slingLineages(SlingshotDataSet(sce))
# specific cluster for KO is marked as a transitory stage.
```
TODO: 
improve plotting.

Separate WT and KO ( cluster specific for KO will not be in the WT, how this affects trajectory? )


```{r eval=FALSE, include=FALSE}
#NOT WORKING
# create a new psuedo time,mean of the two paths
sce$slingPseudotimes <- rowMeans(cbind(sce$slingPseudotime_1, sce$slingPseudotime_2))



# extract information from slingshot object
embed <- embedCurves(sce, "UMAP")
embed <- slingCurves(embed)

# plot umap
umap <- plotUMAP(sce, colour_by = "slingPseudotime_1")
# plot each lineage
for (path in embed) {
embed_l <- data.frame(path$s[path$ord, ])

# i need to figure out how to plot the 
 plot <- umap +
  geom_path(data = embed_l, aes(x = Dim.1, y = Dim.2), size = 1.2)
}

plot
# lineages
slingLineages(SlingshotDataSet(sce))
```

```{r}
# From OSCA
# Need to loop over the paths and add each one separately.
gg <- plotUMAP(sce, colour_by="slingPseudotimes")
embedded <- embedCurves(sce, "UMAP")
embedded <- slingCurves(embedded)
for (path in embedded) {
    embedded_l <- data.frame(path$s[path$ord,])
    gg <- gg + geom_path(data=embedded_l, aes(x=Dim.1, y=Dim.2), size=1.2)
}

gg
```
Ok, the plottings are terrible, but the information is there. 

# Slingshot on the UMAP space.

This is the way to go to get nice embedded lines in the plot. 

```{r}
# run slingshot
sce_umap <- slingshot(sce, clusterLabels = "clusters_named", reducedDim= "UMAP", start.clus="OPC")
summary(sce$slingPseudotime_1)
plotReducedDim(sce_umap, "UMAP", colour_by = "slingPseudotime_1") 

cols <- pals::viridis(100)[cut(sce$slingPseudotime_1, breaks = 100)]

plot(reducedDims(sce)$UMAP, col = cols)
lines(SlingshotDataSet(sce_umap), lwd=2, col='black', type = "lineages")

#plotReducedDim(sce, "corrected", colour_by= "clusters_named")
```

Just... no. No. iOligos should be in the path. 

I tried another seed for the umap, where the oligo cluster is upside-down, and then the starting cluster is suddenly oligo-3 and then it branches to oligo4, oligo2 and oligo1. 